### 2014-04-09 patchy-selection-paper.tex
debugged exponents!

Migration--selection balance ensures that 
there will be some number of $B$ alleles present in the regions where they are disadvantageous,
but only very few, and rarely, far away from the patch where $B$ is established.
Denote the expected frequency of allele $B$ at displacement $x$ relative to the patch by $q(x)$.
Following \citet{slatkin1973geneflow}, one can write down a differential equation to which $q(x)$ is the solution,
which in one dimension can be solved to give $q(x)$ in terms of Jacobi elliptic functions (see Appendix \ref{apx:elliptic_integrals}).
More general, asymptotics of the solution implies that for large $r$ the mean occupation frequency decays exponentially, with a square-root correction in two dimensions (see Appendix \ref{apx:asymptotics}): 
\begin{align} \label{eqn:eqfreq}
    q(x) \approx C (|x|-r_0)^{-(d-1)/2} \exp( - |x| \sqrt{2 s_m} / \sigma) \qquad \text{for large $|x|$},
\end{align}
where $d$ is the dimension ($d=1$ or $2$), and $C$ and $r_0$ are constants depending on the geographic shape of the populations and the selection coefficients
(but note that $r_0$ is only necessary in $d=2$).
In applications we fit $C$ and $r_0$ to the data;
to get concrete numbers from theory we take $C=1$ and $r_0=0$ if necessary.
From simulations, $C$ seems to be close to 1 in practice (see Appendix \ref{apx:patchy_sims})
and XXX $r_0$.

@@

\begin{figure}[ht!]
  \begin{center}
    \includegraphics{sim-occupation-freqs}
  \end{center}
  \label{fig:sim_occupation_freqs}
  \caption{
  Mean occupation frequencies from simulation; details are as in figure~\ref{fig:sim_snapshots} except as otherwise noted.
  Each panel shows mean occupation frequencies of individual demes (grey dots)
  and averaged over demes at similar distances (black circles), on
  \textbf{(A)} a one-dimensional grid of 150 demes, and
  \textbf{(B)} a two-dimensional grid of $50\times 50$ demes.
  Superimposed in black is the decay predicted by \eqref{eqn:eqfreq}.
  The two distinct lines of grey points in (A) are to the left and right of the patch, respectively.
  }
\end{figure}

@@

First, we compute the second term in equation \eqref{eqn:gestalt_q}, (mean occupation density of a family at $x$),
by which we mean the expected total number of offspring of a single family that ever live at distance $r$ from the patch.
In two dimensions, this depends on where the family began, 
and right-hand side of equation \eqref{eqn:gestalt_q} should be integrated over this starting point.
Here we'll write $x_0$ for the starting point, and $u(x_0,x)$ for the mean occupation density at $x$.
Concretely, suppose that $Z$ is a spatial branching process as above, started with a single individual at $x_0$,
defined now so that $Z_t(A)$ is the number of descendant individuals alive at time $t$ living in the set $A$.
The mean occupation density of $Z$ is then defined to be the density of $\E[\int_0^t Z_t dt]$,
i.e.\ the smooth function $u(x_0,x)$ that, integrated over a set $A$, 
gives the expected total number of individuals ever living in $A$:
\begin{align} \label{eqn:occupation_density_defn}
    \E\left[ \int_0^\infty Z_t(A) dt \right] = \int_A u(x_0,x) dx .
\end{align}
Since a family has on average $(1-s_m)^t$ members in generation $t$,
and the marginal distribution of any one surviving individual is approximately Gaussian with variance $t \sigma^2$,
exchanging the expectation and the integral shows that
the density $u(x_0,x)$ is equal to the Green's function for Brownian motion killed at rate $1-s_m$,
\begin{align} \label{eqn:occupation_density_solution}
    u(x_0,x) = \int_0^\infty (1-s_m)^t p_{\sigma^2 t}(x_0,x) dt
\end{align}
where $p_t(x,y) = (2 \pi t)^{-d/2} \exp(-\|x-y\|^2/2t)$ is the transition density for standard Brownian motion.
The integral can be expressed in terms of modified Bessel functions:
using \citet{gradshteyn2007table} (3.471.9 and 8.451.6), the mean occupation density is,
with $\alpha = -\log(1-s_m)$ and $d$ the dimension,
\begin{align}
    u(x_0,x) &= \frac{1}{\sigma^d \pi}\left(\sqrt{ \frac{\|x-x_0\|^2}{2\alpha\sigma^2} }\right)^{(d+2)/2} K_{(d+2)/2}\left( \sqrt{2\alpha} \frac{\|x-x_0\|}{\sigma} \right) \\
             &= \frac{1}{\sigma^d \sqrt{2\pi}} \left(\sqrt{ \frac{\|x-x_0\|^2}{2\alpha\sigma^2} }\right)^{d/2} \exp\left(- \sqrt{2\alpha} \frac{\|x-x_0\|}{\sigma} \right)+ O(\sigma/\|x-x_0\|) . \label{eqn:occupation_asymptotics}
\end{align}
Since we only have asymptotics for the equilibrium frequency, 
we will only use the latter (asymptotic) form.

@@

If $q(r)$ is a radially symmetric equilibrium frequency and $\xi(t,x)=q(|x|)$,
then $\partial_t \xi(t,x) = 0$.
So, using \eqref{eqn:radial_fkpp}, a radially symmetric solution 
with $\gamma(r) = -s < 0$ for all $r>r_0$ 
solves, for $r_0 < r < \infty$,
\begin{align}
    \partial_r^2 q(r) + \frac{d-1}{r} \partial_r q(r) - \frac{2s}{\sigma^2} q(r) (1-q(r)) = 0 \\
    \lim_{r \to \infty} q(r) = \lim_{r \to \infty} \partial_r q(r) = 0 \qquad 
    0 < q(r) < 1 
\end{align}
Since $q(r) \to 0$ as $r \to \infty$, so $q(r) (1-q(r)) \approx q(r)$,
XXX theory in \citet{wasow1965asymptotic} XXX
implies that the true equilibrium frequency $q$ is close, for large $r$, to the solution to
\begin{align}
    \partial_r^2 u(r) + \frac{d-1}{r} \partial_r u(r) - \frac{2s}{\sigma^2} u(r) = 0  \label{eqn:bessel} \\
    \lim_{r \to \infty} u(r) = \lim_{r \to \infty} \partial_r u(r) = 0 \qquad  .
\end{align}
This has general solution given by a modified Bessel function:
using \citet{gradshteyn2007table} 8.494.8,
the general solution is
\begin{align}
    u(r) = C' (r-r_0)^{d/2} K_{d/2} \left( (r-r_0) \sqrt{2s}/\sigma \right) ,
\end{align}
where $C'$ and $r_0$ are chosen to match boundary conditions.
Asymptotics of Bessel functions \citep[][8.451.6]{gradshteyn2007table} then imply that
\begin{align}
    q(r) \simeq C (r-r_0)^{(d-1)/2} \exp \left( -r \sqrt{2s}/\sigma \right) ,
\end{align}
where $C$ is a different constant.

>>>>

Migration--selection balance ensures that 
there will be some number of $B$ alleles present in the regions where they are disadvantageous,
but only very few, and rarely, far away from the patch where $B$ is established.
Denote the expected frequency of allele $B$ at displacement $x$ relative to the patch by $q(x)$.
Following \citet{slatkin1973geneflow}, one can write down a differential equation to which $q(x)$ is the solution,
which in one dimension can be solved to give $q(x)$ in terms of Jacobi elliptic functions (see Appendix \ref{apx:elliptic_integrals}).
More general, asymptotics of the solution implies that for large $|x|$ the mean occupation frequency decays exponentially, with a square-root correction in two dimensions (see Appendix \ref{apx:asymptotics}): 
\begin{align} \label{eqn:eqfreq}
    q(x) \approx C |x|^{-(d-1)/2} \exp( - |x| \sqrt{2 s_m} / \sigma) \qquad \text{for large $|x|$},
\end{align}
where $d$ is the dimension ($d=1$ or $2$), and $C$ and $r_0$ are constants depending on the geographic shape of the populations and the selection coefficients
(but note that $r_0$ is only necessary in $d=2$).
In applications we fit $C$ and $r_0$ to the data;
to get concrete numbers from theory we take $C=1$ and $r_0=0$ if necessary.
From simulations, $C$ seems to be not too far from 1 in practice (see Appendix \ref{apx:patchy_sims}).

@@

\begin{figure}[ht!]
  \begin{center}
    \includegraphics{sim-occupation-freqs}
  \end{center}
  \label{fig:sim_occupation_freqs}
  \caption{
  Mean occupation frequencies from simulation; details are as in figure~\ref{fig:sim_snapshots} except as otherwise noted.
  Each panel shows mean occupation frequencies of individual demes (grey dots)
  and averaged over demes at similar distances (black circles), on
  \textbf{(A)} a one-dimensional grid of 201 demes, and
  \textbf{(B)} a two-dimensional grid of $101\times 101$ demes.
  Superimposed in black is the decay predicted by \eqref{eqn:eqfreq},
  with $C$ chosen arbitrarily to give a reasonable fit.
  }
\end{figure}

@@

First, we compute the second term in equation \eqref{eqn:gestalt_q}, (mean occupation density of a family at $x$),
by which we mean the expected total number of offspring of a single family that ever live at distance $r$ from the patch.
In two dimensions, this depends on where the family began, 
and right-hand side of equation \eqref{eqn:gestalt_q} should be integrated over this starting point.
Here we'll write $x_0$ for the starting point, and $u(x_0,x)$ for the mean occupation density at $x$.
Concretely, suppose that $Z$ is a spatial branching process as above, started with a single individual at $x_0$,
defined now so that $Z_t(A)$ is the number of descendant individuals alive at time $t$ living in the set $A$.
The mean occupation density of $Z$ is then defined to be the density of $\E[\int_0^t Z_t dt]$,
i.e.\ the smooth function $u(x_0,x)$ that, integrated over a set $A$, 
gives the expected total number of individuals ever living in $A$:
\begin{align} \label{eqn:occupation_density_defn}
    \E\left[ \int_0^\infty Z_t(A) dt \right] = \int_A u(x_0,x) dx .
\end{align}
Since a family has on average $(1-s_m)^t$ members in generation $t$,
and the marginal distribution of any one surviving individual is approximately Gaussian with variance $t \sigma^2$,
exchanging the expectation and the integral shows that
the density $u(x_0,x)$ is equal to the Green's function for Brownian motion killed at rate $1-s_m$,
\begin{align} \label{eqn:occupation_density_solution}
    u(x_0,x) = \int_0^\infty (1-s_m)^t p_{\sigma^2 t}(x_0,x) dt
\end{align}
where $p_t(x,y) = (2 \pi t)^{-d/2} \exp(-\|x-y\|^2/2t)$ is the transition density for standard Brownian motion.
The integral can be expressed in terms of modified Bessel functions:
using \citet{gradshteyn2007table} (3.471.9 and 8.451.6), the mean occupation density is,
with $\alpha = -\log(1-s_m)$ and $d$ the dimension,
\begin{align}
u(x_0,x) &= \frac{\alpha^{(d-2)/2}}{\sigma^d \pi^{d/2}}\left(\frac{ \sqrt{2 \alpha} \|x-x_0\|}{\sigma} \right)^{(2-d)/2} K_{(2-d)/2}\left( \sqrt{2\alpha} \frac{\|x-x_0\|}{\sigma} \right) \\
         &= \frac{\alpha^{(d-2)/2}}{\sigma^d 2 \pi^{(d-1)/2}} \left( \frac{\sqrt{2 \alpha} \|x-x_0\|}{\sigma} \right)^{(1-d)/2} \exp\left(- \sqrt{2\alpha} \frac{\|x-x_0\|}{\sigma} \right)+ O(\sigma/\|x-x_0\|) . \label{eqn:occupation_asymptotics}
\end{align}
Since we only have asymptotics for the equilibrium frequency, 
we will only use the latter (asymptotic) form.

@@

If $q(r)$ is a radially symmetric equilibrium frequency and $\xi(t,x)=q(|x|)$,
then $\partial_t \xi(t,x) = 0$.
So, using \eqref{eqn:radial_fkpp}, a radially symmetric solution 
with $\gamma(r) = -s < 0$ for all $r>r_0$ 
solves, for $r_0 < r < \infty$,
\begin{align}
    \partial_r^2 q(r) + \frac{d-1}{r} \partial_r q(r) - \frac{2s}{\sigma^2} q(r) (1-q(r)) = 0 \\
    \lim_{r \to \infty} q(r) = \lim_{r \to \infty} \partial_r q(r) = 0 \qquad 
    0 < q(r) < 1 
\end{align}
Since $q(r) \to 0$ as $r \to \infty$, so $q(r) (1-q(r)) \approx q(r)$,
XXX theory in \citet{wasow1965asymptotic} XXX
implies that the true equilibrium frequency $q$ is close, for large $r$, to the solution to
\begin{align}
    \partial_r^2 u(r) + \frac{d-1}{r} \partial_r u(r) - \frac{2s}{\sigma^2} u(r) = 0  \label{eqn:bessel} \\
    \lim_{r \to \infty} u(r) = \lim_{r \to \infty} \partial_r u(r) = 0 \qquad  .
\end{align}
This has general solution given by a modified Bessel function:
using \citet{gradshteyn2007table} 8.494.9,
the general solution is
\begin{align}
    u(r) = C' (r-r_0)^{(2-d)/2} K_{(2-d)/2} \left( (r-r_0) \sqrt{2s}/\sigma \right) ,
\end{align}
where $C'$ and $r_0$ are chosen to match boundary conditions.
Asymptotics of Bessel functions \citep[][8.451.6]{gradshteyn2007table} then imply that
\begin{align}
    q(r) \approx C r^{(1-d)/2} \exp \left( -r \sqrt{2s}/\sigma \right) + O(1/r),
\end{align}
where $C$ is a different constant.
